---
title: "Credit Card Fraud Detection with Random Forest - Machine Learning Project"
author: "Ryan Navarro, BSN, RN, CPAN, CCRN Alumnus"
date: "2024-12-03"
output:
  pdf_document: default
  html_document:
    df_print: paged
---



# 1 - GET THE DATA

## Load libraries

```{r}
library(dplyr) 
library(ggplot2) 
library(caret) 
library(ROCR)
library(randomForest) 
library(e1071)
```

## Load the dataset

```{r}
cc <- read.csv('/Users/rn/Desktop/Projects/CC Fraud Detection in R/creditcard.csv')
```



# 2 - EXPLORE THE DATA

## View first few rows

```{r}
head(cc)
```

## Summary statistics

```{r}
summary(cc)
```

## Check structure

```{r}
str(cc)
```

## Check Class for imbalance

### non-fraud = 0, fraud = 1

#### *Imbalance common due to rarity of fraudulent vs legitimate transactions.*

```{r}
table(cc$Class)
```

## Check for missing values

```{r}
sum(is.na(cc))
```
## Visualize Class distribution

```{r}
ggplot(cc, aes(x = factor(Class))) +
  geom_bar(fill = c('blue', 'red')) +
  labs(title = 'Class distribution', x = 'Class', y = 'Count')
```

## Correlation Heatmap (to look at relatioships of Features)

```{r}
library(corrplot)
corr_matrix <- cor(cc %>% select(-Class))
corrplot(corr_matrix, method = 'circle')
```



# 3 - PREPARE THE DATA

## Normalize numerical features

```{r}
cc$Amount <- scale(cc$Amount)
```

## Split data into Features (X) and Target (y)

```{r}
X <- cc %>% select(-Class)
y <- cc$Class
```

## Split data into Training (80%) and Test (20%) Sets

```{r}
set.seed(42)
train_cc <- createDataPartition(y, p=0.8, list=FALSE)
X_train <- X[train_cc, ]
X_test <- X[-train_cc, ]
y_train <- y[train_cc]
y_test <- y[-train_cc]
```

## Check Class for imbalance

```{r}
table(y_train)
```

# 4 - SHORTLIST PROMISING MODELS

## Train Random Forest model using the Training data

```{r}
rf_mod <- randomForest(x = X_train, y = as.factor(y_train), ntree = 100, mtry = 5, importance = TRUE)
```

## Summary of the Random Forest model

```{r}
summary(rf_mod)
```

## Evaluate Feature importance

```{r}
varImp(rf_mod)
```


# 5 - MEASURE PERFORMANCE

## Prediction on Test set
```{r}
rf_predict <- predict(rf_mod, X_test)
```

## Confusion Matrix

```{r}
confusionMatrix(as.factor(rf_predict), as.factor(y_test))
```

## ROC Curve

```{r}
rf_prob <- predict(rf_mod, X_test, type = 'prob')[, 2]
pred <- prediction(rf_prob, y_test)
perf <- performance(pred, 'tpr', 'fpr')
plot(perf, col = 'blue', main = 'ROC Curve')
```



# 6 - DEPLOY THE MODEL

## Save the Trained model for future use.

```{r}
saveRDS(rf_mod, 'CCFraudDetection_rf_mod.rds')
```

## Load the Trained model for reuse
### Eg. predictions <- predict(loaded_model, X_new) 
#### *where X_new is your new data (new features) for prediction

```{r}
loaded_model <- readRDS('CCFraudDetection_rf_mod.rds')
```

